Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-08-01 treestats_paper.tex
comparison of tsinferish things

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them.
To evaluate the correspondence between site and branch statistics
in real data,
we calculated mean sequence diversity along chromosome 20 from the 1000 Genomes dataset
in windows of 500Kb \plr{check this}
separately in each of its 26 populations,
calculated using the site statistic described in Example~\ref{ex:site_diversity},
and compared this to the corresponding branch statistic (of Example~\ref{ex:branch_diversity})
as calculated from three inferred tree sequences:
by tsinfer \citep{tsinfer}, which does not attenpt to estimate well-calibrated node times,
(2) by GEVA \citep{geva}, which aims to calibrate the node times reported by tsinfer,
and (3) by Relate \citep{relate}, an independently developed method.
These are shown in Figure~\ref{fig:chr20}.

@@

These are different because of XXX but similar YYY.
As discussed above, we expect site and branch statistics to agree up to noise
if mutation rates are constant along the genome.
However, this caveat may well be substantial -- mutation rates are likely \emph{not} constant
\citep{mut_rate_variation},
although the extent of this is unknown.

>>>>

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them.
To evaluate the correspondence between site and branch statistics
in real data,
we calculated mean sequence diversity along chromosome 20 from the 1000 Genomes dataset
in windows of 500Kb \plr{check this}
separately in each of its 26 populations,
calculated using the site statistic described in Example~\ref{ex:site_diversity},
and compared this to the corresponding branch statistic (of Example~\ref{ex:branch_diversity})
as calculated from three inferred tree sequences:
by tsinfer \citep{tsinfer}, which does not attenpt to estimate well-calibrated node times,
(2) by GEVA \citep{geva}, which aims to calibrate the node times reported by tsinfer,
and (3) by Relate \citep{relate}, an independently developed method.

The comparisons are shown in Figure~\ref{fig:chr20}.
Of these methods, branch diversity from tsinfer shows the worst agreement with site diversity,
which is unsurprising as tsinfer makes no attempt to provide well-calibrated node times,
and only aims to provide consistent tree topologies.
The two methods that infer node times differ substantially,
with Relate showing a significantly tighter agreement than GEVA.
Does this mean that Relate is doing a better job at inferring the tree sequence than GEVA?
This is highly suggestive, but not entirely clear.

@@

In practice, we do not expect branch and site diversity to agree exactly,
because of local differences in mutation rate and intensities of selected mutations.
For instance, regions in which a high proportion of potential mutations are deleterious
are expected to have a diversity for two reasons:
first, the deleterious mutations themselves are less likely to be found,
and second, the indirect effect of deleterious mutations on nearby sites reduces typical tree height
and thus diversity at even neutral, linked sites \citep{background_selection}.
The first effect would cause site and branch statistics to differ,
because it effectively reduces the mutation rate in the region
and violates the assumption of independence of mutations given the trees.
However, the second effect does not affect the correspondence between site and branch statistics,
since it is mediated by tree shape.
However, it is generally unknown how much mutation rate varies along the genome
\citep[but see][]{mut_rate_varies}
or how dense are targets of selection \citep{how_important_is_selection}.
For this reason, it is very interesting to see how close a correspondence between site and branch
statistics it is \emph{possible} to obtain --
it is tempting to say that the best tree sequence would obtain
as tight a match between site and branch statistics as possible,
with remaining variation explained by selection and/or mutation rate variation.
